---
title: "Title"
subtitle: "Next Generation Sequencing 2025/2026"
author: "Cengiz Can Guner and Matija Zizek"
date: "`r Sys.Date()`"
output: 
  pdf_document: 
    toc: true
    fig_height: 6
    number_sections: true
  html_document: 
    toc: true
editor_options: 
  markdown: 
    wrap: 80
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage 

# Setting The Environment

```{r, message=FALSE, warning=FALSE}
# ============================================================
# Install and Call Required NameSpaces
# ============================================================

# Install BioConductor package if not installed
if (!require("BiocManager", quietly = TRUE)) 
  install.packages("BiocManager")

# Package Lists
cran_pkgs <- c("tidyverse", "corrplot", "ggfortify", "ggplot2")
bioc_pkgs <- c("edgeR", "EDASeq", "limma")

# Check, Install, and Call NameSpaces using CRAN
for (pkg in cran_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
    library(pkg, character.only = TRUE)
  } else {
    library(pkg, character.only = TRUE)
  }
}

# Check, Install, and Call NameSpaces using BioConductor
for (pkg in bioc_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    BiocManager::install(pkg)
    library(pkg, character.only = TRUE)
  } else {
    library(pkg, character.only = TRUE)
  }
}

results_dir <- "results/"
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
```
\newpage

# Explarotary Analysis

```{r, cache=TRUE}
# ============================================================
# Load the Dataser from rds file
# ============================================================
rse <- readRDS("./opt/omicsdata/datasets/SRP029880.rds")
rse
```

The given dataset contains `r dim(rse)[1]` genes across `r dim(rse)[2]` samples.  



```{r}
# ============================================================
# Inspect Data More
# ============================================================
# Inspect row and column names of the data
message("Row names: ", paste(rownames(rse)[1:3], collapse = ", "),
        "\nColumn names: ", paste(colnames(rse)[1:5], collapse = ", "))

# Inspect Sample (colData) and Genes metadata (rowData)
message("Sample metadata: ") 
print(colData(rse)[1:5, 1:5])
message("\nGene metadata: ")
print(rowData(rse)[1:5, ])

```

The column names represent sample names and the row names represent gene IDs.  

The sample metadata contains `r dim(colData(rse))[2]` variables for each `r dim(colData(rse))[1]` sample.

The gene metadata contains `r dim(rowData(rse))[2]` variables for each `r dim(rowData(rse))[1]` genes.

```{r}
# ============================================================
# Extract information from attributes data to a new column
# ============================================================
rse@colData$source_name <- sapply(
  strsplit(rse@colData$sra.sample_attributes, split = "|", fixed = TRUE),
  function(x) strsplit(x[2], split = ";;", fixed = TRUE)[[1]][2]
)
```

```{r, warning=FALSE}
# ============================================================
# Exploratory Plots
# ============================================================
raw_counts  <- assay(rse, "raw_counts")
source_name <- rse$source_name
names(source_name) <- colnames(rse)
log1p_counts <- log1p(raw_counts)

df_long <- as.data.frame(log1p_counts) %>%
  mutate(genes = row.names(.)) %>%
  pivot_longer(cols = -genes, names_to = "samples", values_to = "expression")
df_long$source_name <- source_name[df_long$samples]

ggplot(df_long, aes(x = samples, y = expression, fill = source_name)) +
  geom_boxplot() +
  scale_y_continuous(limits = c(0, 20)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Raw Counts", x = "Samples", y = "log1p Expression", fill = "Source")
```

```{r, cache=TRUE}
exp_plot_dir  <- file.path(results_dir, "exploratory")
ma_plot_dir   <- file.path(exp_plot_dir, "ma_plots")
md_plot_dir   <- file.path(exp_plot_dir, "md_plots")
dir.create(exp_plot_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(ma_plot_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(md_plot_dir, showWarnings = FALSE, recursive = TRUE)


combinations <- combn(c(1:ncol(rse)), 2)
sample_names <- colnames(rse)

# MA Plot
begin <- proc.time()
for (i in 1:ncol(rse)) {
  filename <- paste0("MAplot_", sample_names[i], ".png")
  
  png(filename = file.path(ma_plot_dir, filename), width = 800, height = 600)
  
  limma::plotMA(log1p_counts, array = i, 
                main = paste0(sample_names[i]), xlab = "A", ylab = "M")
  
  dev.off()
}
run_time <- proc.time() - begin
message("Time elapsed during MA Plotting: ", run_time[3], " seconds.")

# MD Plot
begin <- proc.time()
for (i in 1:ncol(combinations)) {
  pair <- combinations[, i]
  
  filename <- paste0("MDplot_", sample_names[pair[1]], "_vs_", sample_names[pair[2]], ".png")
  
  png(filename = file.path(md_plot_dir, filename), width = 800, height = 600)
  
  EDASeq::MDPlot(raw_counts, c(pair[1], pair[2]))
  title(main = paste0(sample_names[pair[1]], " vs ", sample_names[pair[2]]))
  
  dev.off()
}

run_time <- proc.time() - begin
message("Time elapsed during MD Plotting: ", run_time[3], " seconds.")
```

# Normalization

# Differential Expression Analysis

# Gene Enrichment Analysis